\documentclass[11pt]{amsart}
%prepared in AMSLaTeX, under LaTeX2e
\addtolength{\oddsidemargin}{-.9in} 
\addtolength{\evensidemargin}{-.9in}
\addtolength{\topmargin}{-.9in}
\addtolength{\textwidth}{1.7in}
\addtolength{\textheight}{1.5in}

\renewcommand{\baselinestretch}{1.05}

\usepackage{verbatim} % for "comment" environment

\usepackage{palatino}

\usepackage[final]{graphicx}

\usepackage{tikz}
\usetikzlibrary{positioning}

\usepackage{enumitem,xspace,fancyvrb}

\newtheorem*{thm}{Theorem}
\newtheorem*{defn}{Definition}
\newtheorem*{example}{Example}
\newtheorem*{problem}{Problem}
\newtheorem*{remark}{Remark}

\DefineVerbatimEnvironment{mVerb}{Verbatim}{numbersep=2mm,frame=lines,framerule=0.1mm,framesep=2mm,xleftmargin=4mm,fontsize=\footnotesize}

% macros
\usepackage{amssymb}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\bE}{\mathbf{E}}
\newcommand{\bF}{\mathbf{F}}
\newcommand{\bJ}{\mathbf{J}}

\newcommand{\bb}{\mathbf{b}}
\newcommand{\bi}{\mathbf{i}}
\newcommand{\bj}{\mathbf{j}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}

\newcommand{\ppr}[1]{\frac{\partial #1}{\partial r}}
\newcommand{\ppt}[1]{\frac{\partial #1}{\partial t}}
\newcommand{\ppx}[1]{\frac{\partial #1}{\partial x}}
\newcommand{\ppy}[1]{\frac{\partial #1}{\partial y}}
\newcommand{\ppz}[1]{\frac{\partial #1}{\partial z}}

\newcommand{\Div}{\ensuremath{\nabla\cdot}}
\newcommand{\Curl}{\ensuremath{\nabla\times}}

\newcommand{\eps}{\epsilon}
\newcommand{\grad}{\nabla}
\newcommand{\ip}[2]{\ensuremath{\left<#1,#2\right>}}
\newcommand{\lam}{\lambda}
\newcommand{\lap}{\triangle}

\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\prob}[1]{\bigskip\noindent\textbf{#1.}\quad }

\newcommand{\Matlab}{\textsc{Matlab}\xspace}

\newcommand{\ds}{\displaystyle}

\begin{document}
\scriptsize \noindent Math 615 NADE (Bueler) \hfill 8 April 2023 \quad \fbox{\emph{Nothing to turn in!}}
\normalsize\medskip

\Large\centerline{\textbf{Von Neumann analysis: plug and chug}}
\medskip
\normalsize

\thispagestyle{empty}
\begin{quote}
\emph{Section 9.6 of the textbook (R.~J.~LeVeque, 2007) presents von Neumann analysis without emphasizing how people actually \emph{do} the analysis.  This worksheet exercises the standard style.\footnote{The textbook eventually says it clearly in section 10.5.}}
\end{quote}


\prob{Example. FTCS on heat equation}  It is easiest to explain the idea relative to an example.  Suppose we apply the FTCS scheme to the heat equation $u_t = D u_{xx}$ with constant diffusivity $D>0$:
\begin{equation}
\frac{U_j^{n+1} - U_j^n}{k} = D \frac{U_{j-1}^n - 2 U_j^n + U_{j+1}^n}{h^2}  \label{ftcs}
\end{equation}
To find what time steps $k>0$ would be stable for a given spacing $h>0$, von Neumann substituted
\begin{equation}
U_j^n = g(\xi)^n e^{ijh\xi}  \label{ansatz}
\end{equation}
into scheme \eqref{ftcs}, where $\xi \in \RR$ is the \emph{wave number} for the spatial wave $e^{ijh\xi}$, in which $i=\sqrt{-1} \in \CC$ (as usual).  The scalar function $g(\xi)$ is  called the \emph{amplification factor} of the scheme.

The spatial wave is complex, but it really is a wave.  In fact, for the interval $0\le x \le 1$ the grid is $x_j = jh$ and thus
    $$e^{ijh\xi} = e^{i \xi x_j} = \cos(\xi x_j) + i \sin(\xi x_j).$$

Now we want to find $g(\xi)$.  To compute it, substitute form \eqref{ansatz} into scheme \eqref{ftcs}.  Indices ``$n+1$,'' ``$j-1$,'' and ``$j+1$'' will turn into powers.  Then use the properties of the exponential.  After simplification and trigonometric identities---do the details in Exercise 1---we get
    $$g(\xi) = 1 - \frac{4Dk}{h^2} \sin^2\left(\frac{\xi h}{2}\right).$$

Absolute stability $|U_j^{n+1}| \le |U_j^{n}|$ corresponds to $|g(\xi)|\le 1$ for all $\xi \in \RR$.  For this scheme\footnote{The same condition is derived via MOL in section 9.3, and then by von Neumann analysis in section 9.6.} we get the condition $\displaystyle k \le \frac{h^2}{2D}$, which tells us that the time step $k$ must be very small if the spacing $h$ is small.

\bigskip
\prob{Exercise 1. FTCS on heat equation}  Label the stencil.  Then fill in the above details.

\bigskip
\hfill \input{tikz/ftcsstencil.tex}
\vfill


\clearpage
\newpage
\prob{Exercise 2. Crank-Nicolson on heat equation}  Label the stencil.  State the scheme.  Do the analysis.

\bigskip
\hfill \input{tikz/cnstencil.tex}
\vfill

\prob{Exercise 3. FTCS on the advection equation $u_t + a u_x = 0$ with $a$ constant}  Write down the scheme, draw and label a stencil, and do the analysis.
\vfill

\prob{Exercise 4. CTCS (leapfrog) on advection equation}  Again.
\vfill





\end{document}
