\documentclass[12pt]{amsart}
\addtolength{\topmargin}{-0.6in} % usually -0.25in
\addtolength{\textheight}{1.1in} % usually 1.25in
\addtolength{\oddsidemargin}{-0.7in}
\addtolength{\evensidemargin}{-0.7in}
\addtolength{\textwidth}{1.5in} %\setlength{\parindent}{0pt}

\newcommand{\normalspacing}{\renewcommand{\baselinestretch}{1.04}\tiny\normalsize}

\normalspacing

% macros
\usepackage{amssymb,xspace}
\usepackage{tikz}
\usepackage[pdftex,colorlinks=true]{hyperref}


\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}

\newcommand{\mtt}{\texttt}
\newcommand{\mtl}[1]{{\texttt{>>#1}}}
\usepackage{alltt}
\usepackage{fancyvrb}

\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}

\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\ZZn}{{\mathbb{Z}}_n}
\newcommand{\NN}{{\mathbb{N}}}

\newcommand{\eps}{\epsilon}
\newcommand{\grad}{\nabla}
\newcommand{\lam}{\lambda}
\newcommand{\ip}[2]{\mathrm{\left<#1,#2\right>}}
\newcommand{\erf}{\operatorname{erf}}

\renewcommand{\Re}{\operatorname{Re}}
\renewcommand{\Im}{\operatorname{Im}}
\newcommand{\Arg}{\operatorname{Arg}}

\newcommand{\Span}{\operatorname{span}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\range}{\operatorname{range}}
\newcommand{\trace}{\operatorname{tr}}
\newcommand{\Null}{\operatorname{null}}

\newcommand{\Matlab}{\textsc{Matlab}\xspace}
\newcommand{\Octave}{\textsc{Octave}\xspace}
\newcommand{\pylab}{\textsc{pylab}\xspace}
\newcommand{\longMOP}{\textsc{Matlab}\big|\textsc{Octave}\big|\textsc{pylab}\xspace}
\newcommand{\MOP}{\textsc{M}\big|\textsc{O}\big|\textsc{p}\xspace}

\newcommand{\prob}[1]{\bigskip\noindent\large\textbf{#1.} \normalsize}
\newcommand{\bookprob}[1]{\bigskip\noindent\large\textbf{Exercise #1.} \normalsize}
\newcommand{\probpart}[1]{\smallskip\noindent\textbf{(#1)}\quad }
\newcommand{\aprobpart}[1]{\textbf{(#1)}\quad }


\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}


\begin{document}
\scriptsize \noindent Math 615 NADE (Bueler) %\hfill Thursday, 4 May 2023
\thispagestyle{empty}

\bigskip
\Large\textbf{\centerline{Final Exam: 2 Short Case Summaries}}

\medskip
\large\textbf{\centerline{Thursday, 4 May 2023, 10:15am--12:15pm, Brooks 302}}

\normalsize
\bigskip
The in-class Final Exam will be a bit different, but short.  I would like you to summarize finite difference (FD) numerical analyses of two of the DE problems we have seen during the semester.  That is, I would like you to write two \emph{Case Summaries}.

\medskip
\centerline{\textbf{Please choose one ODE and one PDE problem from the lists at the bottom.}}

\centerline{\textbf{For each one, write a Case Summary on a separate sheet.}}

\smallskip
\textbf{You may not bring notes to the final.}  However, you are \textbf{strongly encouraged to draft and practice your planned Case Summaries.}  Make sure to read relevant sections of the textbook for your topics.  Feel free to get feedback on your drafts from other students or faculty/friends/family/pets or me.  Then think through, in preparation, how you will remember enough detail so as to recreate the Summary during the Final Exam itself.

\smallskip
Each summary should be at most a page.  Equations are important, plus a few (at most three) sketches as needed, but please write complete sentences for the major ideas.  A Summary which is \emph{only} equations or sketches cannot achieve a good grade.

\smallskip
Except for special cases, a Summary should \textbf{state the problem} (DE and initial/boundary conditions), comment on or supply an \textbf{exact solution}, state one or two \textbf{FD discretization}(s), summarize the \textbf{truncation error}, and summarize \textbf{stability}.  Then it should consider and fill-in any relevant additional ideas for the problem and the chosen method; see ``Consider:'' below for the specific problems.  The goal is \emph{not} to prove anything, but to accurately and precisely summarize what we know---or we easily could know---about the problem and the available method(s), especially in advance of writing a code.

\smallskip
For each Summary your goal is to write something like a neat 10 or 15 minute white-board presentation of how you would solve the problem numerically, as you might give to a hypothetical fellow graduate student who has not seen that particular problem but who has taken similar-level classes.  Give an executive summary, and remember to point out the basic things!

\medskip

\newcommand{\ecomment}[1]{ \hfill \mbox{\emph{#1}} }

\renewcommand{\labelenumi}{ODE \circled{\arabic{enumi}}}
\begin{enumerate}
\setlength{\itemsep}{4pt}
\item A general linear 2nd-order ODE BVP. \ecomment{eqn (2.64)}

\noindent Consider: how to solve the linear system.
\item The nonlinear 2nd-order ODE BVP for a pendulum.  \ecomment{eqn (2.77)}

\noindent Consider: Newton's method, how to solve the linear Newton steps.
\item A non-homogeneous, constant-coefficient linear ODE IVP system.  \ecomment{eqn (5.6)}

\noindent Consider: matrix exponentials, compare $\ge 2$ schemes, absolute stability regions.
\end{enumerate}

\renewcommand{\labelenumi}{PDE \circled{\alph{enumi}}}
\begin{enumerate}
\setlength{\itemsep}{4pt}
\item The Poisson problem PDE in 2D. \ecomment{eqn (3.5)}

\noindent Consider: ordering the unknowns, how to solve the linear system.
\item The time-dependent heat equation PDE in 1D. \ecomment{eqn (9.1)}

\noindent Consider: compare $\ge 2$ schemes.
\item The time-dependent scalar advection equation PDE in 1D.  \ecomment{eqn (10.1)}

\noindent Consider: compare $\ge 2$ schemes.
\end{enumerate}

\end{document}

